---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Potato Trade"
slug: "potato-trade"
subtitle: ""
summary: "Diving deeper into the world 'potato economy'"
authors: [admin]
tags: ["R", "R Markdown", "data", "agro"]
categories: ["Agro"]
date: '`r format(Sys.Date())`'
lastmod: '`r format(Sys.Date())`'
featured: false
bibliography: references.bib
draft: true

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: "Smart"
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: ["agro-blog"]
---

In my last post, I visualized potato production by the world market leaders over the past 60 years, with an additional focus on comparing Peru's production to the global top 10.  

This time, I want to dive a bit deeper into the world "potato economy," integrating a trade matrix dataset with the "crops" data I examined last time. This new dataset provides observations of the Export and Import quantities and values for the same range of crops and countries included in the last dataset.  It also provides another level of detail that will really help me out - the "partner country" involved in the trade. In other words, one row in this trade matrix dataset represents an observation of one measure (quantity/value imported/exported) of one crop in one year between two specific countries.

Working with these two datasets means that I will need to use *relational data* methods - processes and functions to combine/join tables using *key* elements that either already exist in the tables or could be added in a series of steps to unite the data.

Fortunately, I downloaded both of my datasets from the FAOSTAT database, which will makes them pretty easy to join using a function built in to the [**dplyr**](https://blog.rstudio.com/2014/01/17/introducing-dplyr/) package for the R language.  You will see this in the [Joining Data] section.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Read in modules
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)
library(scales)
library(rworldmap)
```

## Importing Data

```{r import-trade-data, cache=TRUE, message=FALSE, echo=TRUE, out.width="100%"}
# Read in the data from csv file, store as "raw_data" - not to be overwritten
raw_data <- read_csv("../data/FAO_potato_matrix.csv", guess_max = 5000, local = locale(encoding = "latin1")) %>% as_tibble()
raw_data %>% head(10) %>% kable(caption = "Raw Data from FAO") %>% kable_styling(full_width = T, bootstrap_options = c("responsive"))
```


## Preparing Data

As you can see in the last two columns of the data frame in Table \@ref(tab:import-trade-data), some observations are marked with a flag. Each observation in this dataset is marked with one of the flags listed below to indicate how the data was gathered or calculated.

```{r drop-flags, cache=TRUE}
allFlags <- distinct(raw_data, Flag)
# Show table of flags and descriptions
raw_data %>% select(Flag, `Flag Description`) %>% distinct(Flag, .keep_all = TRUE) %>% kable(caption = "Flags in the Trade Matrix Dataset") %>% kable_styling(full_width = F, position = "center")

# Filter out the flagged observations and drop the Flag and Flag description columns
working_data <- raw_data %>% filter(is.na(Flag)) %>% subset(select = -c(Flag, `Flag Description`))
```

For the purposes of my analysis in this post, I will only use observations that use the "Official Data," filtering out the other types.  

### Tidy Data

Additionally, I want the data I have for the Import and Export quantities and values to be in their own columns in the data frame, rather than different observations. I just need to keep in mind that the quantities are in units of *tons* and the values are in units of *USD $1000*.  Here, I use the `pivot_wider` function from **tidyr** to accomplish this:

```{r pivot-wider, echo=TRUE, message=FALSE, warning=FALSE}
working_data <- working_data %>% pivot_wider(
  names_from = Element, 
  values_from = Value,
  id_cols = c(`Reporter Country Code`, `Reporter Countries`, `Partner Country Code`, `Partner Countries`, `Item Code`, Item, Year)
  )
```

```{r previous-data, cache=TRUE, message=FALSE}
crops_aggregated <- c("Cereals (Rice Milled Eqv)", "Cereals, Total", "Citrus Fruit, Total", "Coarse Grain, Total", "Crops Primary", "Fibre Crops Primary", "Fruit Primary", "Oilcrops", "Pulses, Total", "Roots and Tubers, Total", "Treenuts, Total", "Vegetables Primary")

production_raw_data <- read_csv("../data/Production_Crops_E_All_Data_NOFLAG.csv", guess_max = 5000, local = locale(encoding = "latin1")) %>% filter(Element == "Production" & Item == "Potatoes") %>% drop_na()
allAreas <- distinct(production_raw_data, Area)[[1]]
areasToRemove <- allAreas[225:length(allAreas)]
```

```{r clean-previous-data, cache=TRUE}
production <- production_raw_data %>% pivot_longer(
  cols = starts_with("Y"),
  names_to = "Year",
  names_prefix = "Y",
  values_to = "Value",
  values_drop_na = TRUE
  ) %>% filter(!(Item %in% crops_aggregated) & !(Area %in% areasToRemove) & !startsWith(Area, "China, ")) %>%
  pivot_wider(
    names_from = Element,
    values_from = Value
  )
production$Year <- as.numeric(production$Year)
```


## Joining Data

```{r join-tables, echo=TRUE, message=FALSE, cache=TRUE}
joined_tables <- working_data %>% left_join(production, by = c("Reporter Country Code" = "Area Code","Item Code", "Item", "Year"))
joined_tables <- drop_na(joined_tables)
```

```{r present-joined-data}
joined_tables %>% head(10) %>% kable(caption = "The two tables, joined") %>% kable_styling(bootstrap_options = "responsive")
```

## Exploration

```{r}
showTop10Destinations <- function(country, year) {
  joined_tables %>% filter(`Reporter Countries` == country & Year == year) %>% arrange(desc(`Export Quantity`)) %>% head(10)
}
```

