---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "FAO Intro"
subtitle: "Examining some basic world crops data"
summary: ""
authors: [admin]
tags: ["R", "RMarkdown", "plot", "data","agro"]
categories: ["Agro"]
date: 2020-05-13T16:48:03-05:00
lastmod: 2020-05-13T16:48:03-05:00
featured: true
bibliography: references.bib
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: ["agro-blog"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Read in modules
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)
library(plotly)
library(htmlwidgets)
library(widgetframe)
library(scales)
```


## Welcome!
With this first post of Agro Blog, I'm going to do some basic exploration of a FAO dataset [@FAOSTAT] which includes measures of production, yield, and area harvested for the whole list of crops and countries recorded, with observations ranging from 1961 to 2018.

In this exploration, I will go through my steps of working with this dataset in R, including the cleaning/tidying process, exploratory analysis, and visualization.

For the purposes of this post, I will examine the 2018 data for both the United States and Peru (the two countries where I have explored agriculture).  After cleaning and tidying the data, for each country I will take a look at the top 10 crops by tonnage of production, and compare this with the top 10 crops by yield. For a crop of interest, I will examine how the country stacks up against worldwide production, and how those production numbers have changed over the time period represented by the data.

## Let's get our data ready to use
```{r data-import, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE, tidy=TRUE, cache=TRUE}
raw_data <- read_csv("../data/Production_Crops_E_All_Data_NOFLAG.csv",guess_max = 5000, local = locale(encoding = "latin1")) %>% as_tibble()
head(raw_data, 10) %>% kable(caption = "Raw dataset from FAOSTAT") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE, position = "center")
```

This data is not ready to use just yet - all of the years are included as column names (and with that messy "Y").  To fix this, I will use a *pivot* to bring all of those years into a single column "Year".

```{r data-cleaning, echo=TRUE, tidy=TRUE, include=TRUE}
crops <- raw_data %>% pivot_longer(
  cols = starts_with("Y"),
  names_to = "Year",
  names_prefix = "Y",
  values_to = "value",
  values_drop_na = TRUE
)
head(crops, 10) %>% kable("html",caption = "Now-tidy Dataset") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE, position = "center")
```

```{r year-to-numeric}
crops$Year <- as.numeric(crops$Year)
```


Now, every row in the dataset represents an observation of one *measure* (production, yield, area harvested) of one crop/crop group in one country for a specific year. This organization is much more *tidy*, and easier to work with.



## Exploratory Data Analysis

```{r define-functions, include=FALSE}
showMostProduced <- function(country, year, number, table_position) {
  crops %>% filter(Area == country & Year == year & Element == "Production") %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Item, value) %>%
    arrange(Rank) %>% head(number) %>% format.data.frame(big.mark=",") %>% kable("html",caption = str_interp("10 Most Produced Crops in ${country}, ${year}"), col.names = c("Rank", "Crop", "Production (tons)")) %>% kable_styling(full_width = FALSE, position = table_position)
}


# Function to graph time series of a measure of one crop for specific country
displayCropInfoByCountry <- function(country, crop, measure) {
  relevantUnit <- crops %>% filter(Element == measure) %>% head(1) %>% select(Unit)
  crops %>% filter(Area == country & Item == crop & Element == measure) %>%
    ggplot(mapping = aes(x = Year, y = value)) + geom_point() + geom_smooth() +
    ggtitle(str_interp("${measure} of ${crop} in ${country}")) + ylab(str_interp("${measure} (${relevantUnit})"))
}

displayCropsInfoByCountries <- function(countries, some_crops, measure) {
  relevantUnit <- crops %>% filter(Element == measure) %>% head(1) %>% select(Unit)
  crops %>% filter(Area %in% countries & Item %in% some_crops & Element == measure) %>%
    ggplot(mapping = aes(x = Year, y = value, color = Area)) + geom_point() + geom_smooth() +
    facet_wrap(~ Item, nrow = 1) +
    ggtitle(str_interp("${measure} of ${some_crops}")) + ylab(str_interp("${measure} (${relevantUnit})"))
}

```

To streamline my workflow of data visualization, I have defined a few functions to help me present tables and render graphs:

* A function to show the *n* most-produced crops in a given country and year
* A function to graph a measure of one crop in one country over time
* A function to graph a measure of multiple crops in multiple countries over time

### United States
Beginning to examine the data from the United States, let's take a look at the 10 crops with the highest production numbers from 2018:

```{r usa-top10}
showMostProduced("United States of America", 2018, 10, "center")
```

The results of Table \@ref(tab:usa-top10) confirm the fairly common knowledge that US agriculture is dominated by the cereals. However, it shows that the data I'm using is combining some crop aggregates with individual crops.  Rather than downloading another dataset from the FAOSTAT site, I'm just going to filter out the rows that represent an aggregate of crops. Alongside the production ranks, I will also include the top-yielding crops.

```{r remove-aggregates, echo=FALSE, tidy=TRUE}
crops_aggregated <- c("Cereals (Rice Milled Eqv)", "Cereals, Total", "Citrus Fruit, Total", "Coarse Grain, Total", "Crops Primary", "Fibre Crops Primary", "Fruit Primary", "Oilcrops", "Pulses, Total", "Roots and Tubers, Total", "Treenuts, Total", "Vegetables Primary")
crops <- crops %>% filter(!(Item %in% crops_aggregated))
```


```{r usa-production-yield, layout="l-body-outset"}
usa_top10_production <- crops %>% filter(Element == "Production" & Area == "United States of America" & Year == 2018) %>% arrange(desc(value)) %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Item, value) %>% head(10) %>%
  format.data.frame(big.mark=",") %>% kable(caption = "Top 10 Most-Produced Crops in USA, 2018", col.names = c("Rank", "Crop", "Production (tons)")) %>% kable_styling(full_width = F, position = "float_left")

usa_top10_yield <- crops %>% filter(Element == "Yield" & Area == "United States of America" & Year == 2018) %>% arrange(desc(value)) %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Item, value) %>% head(10) %>%
  format.data.frame(big.mark=",") %>% kable(caption = "Top 10 Highest-Yielding Crops in USA, 2018", col.names = c("Rank", "Crop", "Yield (Tons/Hectare)")) %>% kable_styling(full_width = F, position = "left")

usa_top10_production
usa_top10_yield
```


After removing those aggregate values, we get a much more informative list of the top 10 crops in the United States.  Not surprisingly, corn (Maize), soybeans, and wheat take the podium.  Since corn is the USA's top crop, let's take a look at how the USA corn production compares to other top countries in 2018.

```{r clean-countries, include=FALSE}
allAreas <- distinct(crops, Area)[[1]]
areasToRemove <- allAreas[225:length(allAreas)]
crops <- crops %>% filter(!(Area %in% areasToRemove) & !startsWith(Area, "China, "))
top10CornCountries <- filter(crops, Year == 2018 & Element == "Production" & Item == "Maize") %>% arrange(desc(value)) %>% select(Area) %>% head(10) %>% dplyr::pull(var = 1)
```

```{r compare-corn}
crops %>% filter(Item == "Maize" & Year == 2018 & Element == "Production") %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Area, Item, value) %>% arrange(Rank) %>% format.data.frame(big.mark=",") %>%
  head(10) %>% kable("html", align = "c", caption = "Top 10 Producers of Corn, 2018", col.names = c("Rank", "Country", "Crop", "Production (tons)")) %>% kable_styling(full_width = F, bootstrap_options = c("striped", "hover"), position = "center")

```

Now, with this Table \@ref(tab:compare-corn) of the top ten corn-producing countries, let's see how their production has changed since 1961.

```{r corn-graph, message=FALSE, cache=TRUE, warning=FALSE, layout="l-body-outset"}
cornGraph <- displayCropsInfoByCountries(top10CornCountries, c("Maize"), "Production")
cornPlot <- ggplotly(cornGraph, width = 800)
```

```{css, echo=FALSE}
div .figure {
  width: 100%;
}
```

```{r plotly-corn-graph, fig.cap="Corn Production", out.width="100%", class.output="full-width"}
frameWidget(cornPlot)
```

Examining Figure \@ref(fig:plotly-corn-graph) above, we can see that production in both the United States and China has increased dramatically in the last half-century, with Brazil making considerable growth as well.  The parallel growth of the US and China suggests the presence of some oligopolistic conditions in the global corn market - those countries had already been established as production leaders, making it difficult for other countries to inject significant competing production.

This dataset does not include any information on domestic vs. foreign consumption for each crop item - an analysis with that data would better inform my opinion on the economic implications of these production data. It would help me answer questions such as "Do the small corn producers produce less because their current production already satisfies domestic consumption, or because the global commodity market forces them out?"

### Peru
While studying abroad in Peru, I took a course in Agrarian Economics, which dealt primarily with the Peruvian agricultural economy and how it developed over time.  Here, I'll repeat some of the same explorations of crops and production in Peru.

```{r peru-top10}
showMostProduced("Peru", 2018, 10, "float_left")
crops %>% filter(Element == "Yield" & Area == "Peru" & Year == 2018) %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Item, value) %>% arrange(Rank) %>% format.data.frame(big.mark=",") %>% head(10) %>% kable(caption = "10 Highest-Yielding Crops in Peru, 2018", col.names = c("Rank", "Item", "Yield (tons/hectare)")) %>% kable_styling(position = "left", full_width = F)
```

I expected to see the presence of sugar cane and potatoes at the top of this list - these two crops reflect both the Spanish colonial history and indigenous history of the region.  As I experienced living there for a semester, the potato crop is extremely important for diets and nutrition in Peru, across many regions.  First, let's look at the 10 biggest producers of potatoes globally.

```{r potato-countries}
crops %>% filter(Item == "Potatoes" & Year == 2018 & Element == "Production") %>% mutate(Rank = dense_rank(desc(value))) %>% select(Rank, Area, value) %>% format.data.frame(big.mark=",") %>%
  arrange(desc(value)) %>% head(10) %>% kable(align = "c", caption = "Top 10 Producers of Potatoes, 2018", col.names = c("Rank", "Country","Production (tons)")) %>% kable_styling(full_width = F, bootstrap_options = c("striped", "hover"), position = "center")
```

I'm surprised that Peru does not make this top-10 list, given the origins of the potato in the Andean highlands.  I'll add Peru in to this list to compare its production to the global leaders.

```{r potato-countries-with-peru}
crops %>% filter(Item == "Potatoes" & Year == 2018 & Element == "Production") %>% mutate(Rank = dense_rank(desc(value))) %>% filter(Rank <= 10 | Area == "Peru") %>% select(Rank, Area, value) %>% format.data.frame(big.mark=",") %>% arrange(match(Area, "Peru"), Rank) %>% kable(align = "c", caption = "Peru vs. Top 10 Producers of Potatoes, 2018", col.names = c("Rank", "Country", "Production (tons)")) %>% kable_styling(full_width = F, position = "center")
```

As I expected, Peru was not much farther down the list, at 14. Let's do another time-series visualization of how production of potatoes in these countries has developed.

```{r potato-graph, message=FALSE, fig.cap="Potato Production", class.output="full-width", cache=TRUE}
topPotatoCountries <- crops %>% filter(Element == "Production" & Item == "Potatoes" & Year == 2018) %>%
  mutate(Rank = dense_rank(desc(value))) %>% filter(Rank <= 10 | Area == "Peru") %>% select(Area) %>%
  dplyr::pull(var = 1)

potatoPlot <- displayCropsInfoByCountries(topPotatoCountries, c("Potatoes"), "Production") %>% ggplotly(width = 800)
frameWidget(potatoPlot)
```

There is a lot more variation in this graph, compared to the time-series graph of corn production in Figure \@ref(fig:plotly-corn-graph).  China's leadership in this global market is a much more recent development than the dominance of United States corn in the previous graph.  Again, these results leave me desiring consumption data to address the other dimension of this market.  In particular, I'd like to find out whether Peru is a net importer or exporter of potatoes, since the commodity is so valuable to the population.  A FAO report [@FAO-potato] cites the *perishability* and high *transport costs* of potatoes, so I would doubt much significant trade between Peru and countries like China and India.  I will likely pursue further investigations of this topic in future posts.


## References
<div id="refs"></div>
